Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference Imaging Photometric Survey.}

### 2021-04-22 cluster_paper/ms.tex
update plots; update ms

\newcommand{\nautorotdenominator}{1{,}641} % plot_autorotation.py, w/ runid=NGC_2516. For nequal<=0, nclose<=1, LSP>0.08. See /doc/20210205_nequal_and_Prot_vs_color_cleanup.txt
\newcommand{\nautorotnumerator}{987} % plot_autorotation.py, For nequal<=0, nclose<=1, LSP>0.08. See /doc/20210205_nequal_and_Prot_vs_color_cleanup.txt
\newcommand{\nautorotnumeratormatching}{507} % plot_autorotation.py, LSP=SPDM period
\newcommand{\ncompstardenominator}{820} % plot_autorotation.py, w/ runid=compstar_NGC_2516. For nequal<=0, nclose<=1, LSP>0.08. 
\newcommand{\ncompstarnumerator}{365} % plot_autorotation.py
\newcommand{\ncompfrac}{44.5\%} % plot_autorotation.py
\newcommand{\nautovscompstardenominator}{705} % plot_autorotation.py, w/ runid=compstar_NGC_2516. For nequal<=0, nclose<=1, LSP>0.08. 
\newcommand{\nautovscompstarnumerator}{374} % plot_autorotation.py
\newcommand{\nautofrac}{53.1\%} % plot_autorotation.py

@@

We then measured the rotation period from the resulting light curve
using the periodogram implementations in \texttt{astrobase}, namely
the traditional Lomb-Scargle, and the Stellingwerf phase-dispersion
minimization algorithm
\citep{lomb_1976,stellingwerf_period_1978,scargle_studies_1982,stellingwerf_period_2011,bhatti_astrobase_2018}.
We used the CDIPS aperture radius that, based on theoretical
expectations, was expected to give the optimal balance between light
from the target star and background-light \citep{Sullivan_2015}.  This
typically resulted in a circular aperture radius of either 1 or 1.5 pixels.
We recorded the top five periodogram peaks from each method, and their
corresponding powers.  Finally, as a check on crowding, we recorded
the number of stars within the aperture with equal or greater
brightness than the target star, and with brightness within 1.25 and
2.5 TESS magnitudes of the target star.

The resulting rotation periods and periodogram powers are reported,
regardless of detection significance, in Table~\ref{tab:maintable}.
To clean the resulting rotation period measurements, we designed a set
of automated cleaning criteria, which yielded sets we denote
$\mathcal{A}$ and $\mathcal{B}$.  For both sets, we considered light
curves for which the peak Lomb-Scargle periodogram period was below 15
days, the normalized Lomb-Scargle power exceeded 0.08, and for which
no equal-brightness or greater companions were within the aperture.
These limits were chosen after inspecting the light curves visually,
to eliminate low-quality rotation periods while retaining high
completeness.  Beyond requiring that the target star be the brightest
star in the aperture, we also required that at most one companion with
brightness exceeding one tenth of the target star's brightness be
present in the aperture according to the Gaia DR2 source catalog.  For
set $\mathcal{A}$, this yielded \nautorotnumerator\ light curves, out
of \nautorotdenominator\ stars for which rotation periods might
plausibly have been detected.  For set $\mathcal{B}$, we additionally
required that the Lomb-Scargle and Stellingwerf phase-dispersion
methods yielded identical periods to within 10\%.  This yielded
\nautorotnumeratormatching\ light curves.  Boolean flags for each set
are included in Table~\ref{tab:maintable}.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.9\textwidth]{f5a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown versus
  dereddened color for all kinematic \cn\ members with Gaia-ESO
  ($N=452$) or GALAH ($N=107$) spectra available.  The GALAH spectra
  comprise slightly over half of the halo stars, due to the
  non-targeted selection function of that survey.  Field stars (gray
  points) are Keck-HIRES measurements of stars with detected
  transiting planets from the Kepler mission
  \citep{berger_identifying_2018}.  Stars in the Pleiades (125\,Myr)
  with Li detections are from \citet{bouvier_pleiades_lirot_2018};
  Praesepe detections (650\,Myr) are shown in violet
  \citep{soderblom_praesepe_1993}.  Points with ${\rm
  EW}\approx0\,$m\AA\ are non-detections.
  The Li measurements for both core and halo stars in \cn\ are
  consistent with a near-Pleiades age.
  {\bf LGB TODO: the increasing EWs for the Gaia-ESO M dwarfs must be erroneous. What's happening there?}
  \label{fig:lithiumcorehalo}
  }
\end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f5b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium and rotation in NGC\,2516.} Points show all
  kinematic \cn\ members with TESS rotation periods in the Gaia-ESO or
  GALAH surveys, and are colored by equivalent width.  Points with
  reported equivalent widths below 20\,m\AA\ (including
  non-detections) are shown with gray crosses.  The lithium-rotation
  correlation for the K-dwarfs has been observed in other young
  clusters, and is discussed in Section~\ref{subsec:lithium}.
	\label{fig:lithiumrot}
	}
\end{figure*}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.3\textwidth]{f10.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Different clustering techniques yield different
  candidate members of \cn.} \nkinematic\ unique candidate cluster
  members found using three different techniques are considered in
  this study.  \citetalias{cantatgaudin_gaia_2018}:
  \citet{cantatgaudin_gaia_2018},
  \citetalias{kounkel_untangling_2019}:
  \citet{kounkel_untangling_2019}, \citetalias{meingast_2021}:
  \citet{meingast_2021}.
  \label{fig:venn}
	}
\end{figure}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=\textwidth]{f14.pdf}
	\end{center}
	\vspace{-1cm}
  \caption{ {\bf \cn\ light curves sorted in order
  of stellar color.} Fifty stars are randomly selected from Set
  $\mathcal{B}$ (see Section~\ref{subsubsec:cluster}), and the light
  curves are detrended and cleaned as discussed in
  Appendix~\ref{app:detrending}.  Each light curve is separated by
  7.5\% in relative flux.  \label{fig:lightcurves}
	}
\end{figure}

@@

\section{Comparison Star Rotation Periods}
\label{app:compstar}

\begin{figure}[t]
	\begin{center}
		\leavevmode
    \includegraphics[width=0.49\textwidth]{f9.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Rotation in NGC\,2516 compared to the field.}
  Stars with Gaia $Rp<13$\,mag in the cluster and field samples were
  considered. (Fainter field stars were not reduced during
  the initial CDIPS reductions).  The field stars (orange) show a
  different rotation period distribution than the kinematically
  selected \cn\ members.
  Both field and \cn\ stars are subject to the same ``Set $\mathcal{A}$''
  cleaning procedures.
  {\bf LGB TODO: Rerun the field measurements with a better grid to fix the aliasing issue.}
  \label{fig:compstar}
	}
\end{figure}

To provide a basis for comparison, we also opted to search the
``calibration'' light curves ($Rp<13$) that were created as a part of
the CDIPS project.  Over the southern sky (Sectors 1-13 of TESS), this
corresponded to a sample of \ncalibration\ stars.  Cross-matching these
against the \nnbhd\ randomly drawn stars in the neighborhood of \cn yielded
\nnbhdcalibstar\ unique stars, with a cumulative total of
\nnbhdcaliblc\ TESS
sectors observed.
The magnitude cut of $Rp<13$ at the distances of the neighborhood sample
corresponds to an extinction-corrected color cutoff of $\bpmrpo
\approx 0.80$, or spectral types of $\approx$G1V.
This reaches sufficiently far down the main-sequence to enable a
comparison to the cluster star sample.

We performed the same light curve stitching and period-search
procedure on the field comparison stars.
Imposing the same requirements for crowding resulted in
\ncompstardenominator\ stars for which rotation periods could have been
detected.
Imposing the same Lomb Scargle power cutoff, and period upper limit,
yielded \ncompstarnumerator\ period detections (\ncompfrac).
Within the same brightness cutoff, 
\nautovscompstarnumerator\ of \nautovscompstardenominator\ cluster stars
yielded period detections (\nautofrac).
Though the detection fractions are frankly not very different
(likely because of the brightness cutoff), the period {\it
vs} color distributions are quite different
(Figure~\ref{fig:compstar}).

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f12a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.49\textwidth]{f12b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Projection effects for co-moving populations across
  the sky.} The map is colored by the velocity difference a star
  co-moving with \cn\ would exhibit across the equatorial sphere.
  Actual positions of candidate \cn\ members are shown with points;
  the star denotes the cluster center.
	\label{fig:vproj}
	}
\end{figure}

Figure~\ref{fig:vproj} shows why the correct approach for comparing
any star's proper motions against the cluster's mean depends on the
position of the star being considered.

>>>>

\newcommand{\nautorotdenominator}{1{,}641} % plot_autorotation.py, w/ runid=NGC_2516. For nequal<=0, nclose<=1, LSP>0.08. See /doc/20210205_nequal_and_Prot_vs_color_cleanup.txt
\newcommand{\nautorotnumerator}{987} % plot_autorotation.py, For nequal<=0, nclose<=1, LSP>0.08. See /doc/20210205_nequal_and_Prot_vs_color_cleanup.txt
\newcommand{\nautorotnumeratormatching}{507} % plot_autorotation.py, LSP=SPDM period
\newcommand{\ncompstardenominator}{820} % plot_compstar_rotation.py
\newcommand{\ncompstarnumerator}{312} % plot_compstar_rotation.py
\newcommand{\ncompfrac}{38.0\%} % plot_compstar_rotation.py, based on plot_autorotation
\newcommand{\nautovscompstardenominator}{705} % plot_autorotation.py, w/ runid=compstar_NGC_2516. For nequal<=0, nclose<=1, LSP>0.08. 
\newcommand{\nautovscompstarnumerator}{374} % plot_autorotation.py
\newcommand{\nautofrac}{53.1\%} % plot_autorotation.py

@@

We then measured the rotation period from the resulting light curve
using the periodogram implementations in \texttt{astrobase}, namely
the traditional Lomb-Scargle, and the Stellingwerf phase-dispersion
minimization algorithm
\citep{lomb_1976,stellingwerf_period_1978,scargle_studies_1982,stellingwerf_period_2011,bhatti_astrobase_2018}.
We used the CDIPS aperture radius that, based on theoretical
expectations, was expected to give the optimal balance between light
from the target star and background-light \citep{Sullivan_2015}.  This
typically resulted in a circular aperture radius of either 1 or 1.5 pixels.
We recorded the top five periodogram peaks from each method, and their
corresponding powers.  Finally, as a check on crowding, we recorded
the number of stars within the aperture with greater
brightness than the target star, and with brightness within 1.25 and
2.5 TESS magnitudes of the target star.

The resulting rotation periods and periodogram powers are reported,
regardless of detection significance, in Table~\ref{tab:maintable}.
To clean the resulting rotation period measurements, we designed a set
of automated cleaning criteria, which yielded sets we denote
$\mathcal{A}$ and $\mathcal{B}$.  For both sets, we considered light
curves for which the peak Lomb-Scargle periodogram period was below 15
days, the normalized Lomb-Scargle power exceeded 0.08, and for which
no companions with brightness exceeding one-tenth of the target star
were in the aperture ({\it i.e.}, excluding visual binaries
with $\Delta T < 1.25$, according to the Gaia DR2 source catalog).
These limits were chosen after inspecting the light curves visually,
to eliminate low-quality rotation periods while retaining high
completeness.  For set $\mathcal{A}$, this yielded \nautorotnumerator\
light curves, out of \nautorotdenominator\ stars for which rotation
periods might plausibly have been detected.  For set $\mathcal{B}$, we
additionally required that the Lomb-Scargle and Stellingwerf
phase-dispersion methods yielded identical periods to within 10\%.
This yielded \nautorotnumeratormatching\ light curves.  Boolean flags
for each set are included in Table~\ref{tab:maintable}.
A set of field stars was also analyzed for comparison.
The results are discussed in Appendix~\ref{app:compstar}.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.9\textwidth]{f5a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown versus
  dereddened color for all kinematic \cn\ members with Gaia-ESO
  ($N=452$) or GALAH ($N=107$) spectra available.  The GALAH spectra
  comprise slightly over half of the halo stars, due to the
  non-targeted selection function of that survey.  Field stars (gray
  points) are stars with detected
  transiting planets from Kepler 
  \citep{berger_identifying_2018}.  Stars in the Pleiades (125\,Myr)
  with Li detections are from \citet{bouvier_pleiades_lirot_2018};
  Praesepe detections (650\,Myr) are shown in violet
  \citep{soderblom_praesepe_1993}.  Points with ${\rm
  EW}\approx0\,$m\AA\ are non-detections.
  The Li measurements for both core and halo stars in \cn\ are
  consistent with a near-Pleiades age.
  {\bf LGB TODO: the increasing EWs for the Gaia-ESO M dwarfs must be erroneous. What's happening there?}
  \label{fig:lithiumcorehalo}
  }
\end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f5b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium and rotation in NGC\,2516.} Points show all
  kinematic \cn\ members with TESS rotation periods in the Gaia-ESO or
  GALAH surveys, and are colored by equivalent width.  Points with
  reported equivalent widths below 20\,m\AA\ (including
  non-detections) are shown with gray crosses.  The lithium-rotation
  correlation for the K-dwarfs has been observed in other young
  clusters, and is discussed in Section~\ref{subsec:lithium}.
	\label{fig:lithiumrot}
	}
\end{figure*}

@@

\begin{figure}[b]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.3\textwidth]{f10.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Different clustering techniques yield different
  candidate members of \cn.} \nkinematic\ unique candidate cluster
  members found using three different techniques are considered in
  this study.  \citetalias{cantatgaudin_gaia_2018}:
  \citet{cantatgaudin_gaia_2018},
  \citetalias{kounkel_untangling_2019}:
  \citet{kounkel_untangling_2019}, \citetalias{meingast_2021}:
  \citet{meingast_2021}.
  \label{fig:venn}
	}
\end{figure}

@@

\begin{figure}[b]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f14.pdf}
	\end{center}
	\vspace{-1cm}
  \caption{ {\bf \cn\ light curves sorted in order
  of stellar color.} Fifty stars are randomly selected from Set
  $\mathcal{B}$ (see Section~\ref{subsubsec:cluster}), and the light
  curves are detrended and cleaned as discussed in
  Appendix~\ref{app:detrending}.  Each light curve is separated by
  7.5\% in relative flux.  \label{fig:lightcurves}
	}
\end{figure}

@@

\section{Rotation Periods for Comparison Field Stars}
\label{app:compstar}

\begin{figure}[t]
	\begin{center}
		\leavevmode
    \includegraphics[width=0.69\textwidth]{f9.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Rotation in \cn\ compared to the field.}
  Stars with Gaia $Rp<13$\,mag in the cluster and field samples are
  displayed; fainter field stars were not analyzed.  Field
  stars (orange) show a different rotation period distribution than
  the kinematically selected \cn\ members (black).  Field and \cn\
  stars are subject to the same ``Set $\mathcal{A}$''
  period-measurement and cleaning procedures.
  \label{fig:compstar}
	}
\end{figure}

To verify the significance of the trends seen in Figures~\ref{fig:rot}
and~\ref{fig:binarity}, we searched the CDIPS calibration light curves
for rotation periods.  This light curve database comprises all $Rp<13$
stars that fell on TESS silicon. The need to distinguish field and
cluster stars is based on our processing capabilities and is discussed
by \citet{bouma_cdipsI_2019}.  Over the southern sky (Sectors 1-13 of
TESS), this corresponded to a sample of \ncalibration\ stars.
Cross-matching these against the \nnbhd\ randomly drawn stars in the
neighborhood of \cn\ yielded \nnbhdcalibstar\ unique stars.  The
magnitude cut of $Rp<13$ at the distances of the neighborhood sample
corresponds to an extinction-corrected color cutoff of $\bpmrpo
\lesssim 0.80$, or spectral types of $\lesssim$G1V.  This reaches
sufficiently far down the slow sequence to enable a comparison against
the cluster star sample.

We performed the same light curve stitching and period-search
procedure discussed in the text on the field stars.
Imposing the same requirements for crowding resulted in
\ncompstardenominator\ stars for which rotation periods could have
been detected.  Imposing the same Lomb Scargle power cutoff (${\rm
LSP}>0.08$) and period upper limit ($P<15\,{\rm days}$) yielded
\ncompstarnumerator\ period detections (\ncompfrac).  Within the same
brightness cutoff, \nautovscompstarnumerator\ of
\nautovscompstardenominator\ cluster stars yielded period detections
(\nautofrac).  While the detection fractions are only mildly
different (likely due to the brightness cutoff), the period--color
distributions are extremely different
(Figure~\ref{fig:compstar}).
This confirms our expectation that our period measurement procedure is
not biased in any way that could produce the features we observe in
\cn.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f12a.pdf}
			\includegraphics[width=0.49\textwidth]{f12b.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{ {\bf Projection effects for co-moving populations across
  the sky.} The maps are colored by the proper motion difference a star
  co-moving with \cn\ would exhibit across the equatorial sphere.
  Actual positions of candidate \cn\ members are shown with points;
  the star denotes the cluster center.
  The {\it left} and {\it right} panels show the proper motion correction in
  the declination and right ascension directions.
	\label{fig:vproj}
	}
\end{figure}

To compare any given star's proper motion against the cluster's mean,
one must consider the position of the star on the sky.  This is
because two stars sharing identical cartesian velocities will have
different velocities when projected on the sky, and was emphasized by
\citet{meingast_2021}.
Figure~\ref{fig:vproj} shows the magnitude of the correction in
equatorial tangential velocities.
